An agent-based model of military mechanization

### 2018-07-12 paper2/paper2.tex
background

Military mechanization is the degree to which a state's army is comprised of
ground combat vehicles as opposed to infantry. Mechanization is both a 
theoretically important topic in international relations scholarship and a practical 
concern for policymakers. \citet{sechser2010army} influentially argue that
inter-state dynamics (e.g., rivalries) better explain mechanization choices than do
intra-state factors (e.g., culture). This paper demonstrates that
the multivariate statistical approach employed in \citet{sechser2010army} is
	inadequate and that agent-based modeling is needed to capture
	international security competition as a complex system. Candidate
	agent-based models are presented.

@@

@@

Multicollinearity - everything correlated with each other

@@

This paper presents an agent-based model of military mechanization. First, it
provides a theoretical discussion of military mechanization, states the research
hypothesis, and explains why agent-based modeling is an appropriate research strategy.
Then, it details an original agent-based model; presents results; and discusses
their implications for the academic debate on military mechanization.

In summary, the orthodox view of military mechanization is that it is a
reflexive process: states choose their mechanization posture largely based on
that neighbors, enemies, and recent military experiences (e.g. defeat in
counter-insurgency). This paper argues
that this view---based on traditional longitudinal data analysis---is ripe for
agentization. It implicitly argues for autonomous agents (states) that have
cognition regarding mechanization that takes into account other agents (e.g.
enemies), spatiality (e.g. neighbors), and learning from experience. This paper
builds such a model and validates it against the empirical record, using the
National Mechanization Index of \citet{sechser2010army}. It finds that the 
orthodox view of mechanization does not stand up to agentization. A positive feedback loop 
causes states to
mechanize much faster and to a greater degree than occurs in the real-world.
This implies the existence of some countervailing factor (possibly a negative
feedback loop) that mitigates the extent of mechanization. Candidates for what
this could be mostly center on ``internal'' factors---e.g., culture, economy,
and so on---as opposed to the ``external'', security environment-centric
orthodox view.

\section{Background}

\subsection{Military mechanization}

States differ in the composition of their military ground forces.\footnote{States may have multiple organizations that
conduct ground combat operations: e.g., the United States operates both an Army
and a Marine Corps as independent services. Henceforth, this paper uses the
terms ``army'' and ``ground forces'' interchangeably.} One notable axis of comparison 
is \textit{mechanization}: the degree to which an army consists of infantry 
vs. combat vehicles. Treating mechanization as a continuous variable, 
one extreme is an army consisting exclusively of dismounted, small 
arms-wielding infantry units; the other extreme is an army consisting 
solely of tank units, armored personnel carriers, artillery, and the like.
Clearly, no modern state embodies either extreme, but the 1970s-era Vietcong are
a good example of a low-mechanization army and the 1980s-era Israeli army of
high mechanization.

Practical and academic consequences

The consequences of mechanization are large. There is
academic consensus that defense policy---including but not limited to
mechanization---conditions battlefield effectiveness. 
\citet{lyall2009rage} finds that highly mechanized militaries are less effective
at counterinsurgency, while \citet{biddle2004military} argues that
mechanization-enabled mobility aids conventional warfighting. It thus
pays to have an army that is designed to win the
kinds of
conflicts that a state is likely to face. Lacking a crystal ball, however,
leaders cannot always forecast conflict, and armies are large, complex
organizations that typically are held to change only infrequently and slowly
\citep{murray1998military,locher2004victory,zegart2000flawed}. A state suddenly faced with
a war for which its mechanization posture is ill-suited can either concede or
pay a cost in blood and treasure; what it cannot do is
quickly overcome the constraints imposed by past mechanization decisions. 
A quote from then-Secretary of Defense Donald Rumsfeld---made in 2004, when the U.S. military
was confronted in Iraq with the consequences of past choices regarding
mechanization---emphasizes the point:
``you go to war with the army you have'' \citep{schmitt_2004}.

The causes of military mechanization are disputed. \textit{Realists} believe 
that mechanization choices are driven by security environment. States
choose mechanization policies that they believe will allow them to prevail against potential adversaries, or to send a
signal of deterrence, or to learn from the perceived mistakes of the last war
\citep{mearsheimer1983conventional,huth1988extended,murray2011military}. The other school of thought is more
diverse and so harder to name, but may be termed 
\textit{institutionalists}. They believe that factors other than 
strategic calculation determine military mechanization. Domestic political
institutions---e.g., democracy vs. autocracy, politically stable vs.
coup-prone, the state of civil-military relations, and so on---may influence 
force structure \citep{reiter2002democracies,quinlivan1999coup,talmadge2015dictator,brooks2008shaping}. 
Economic factor
endowments may play a role: e.g., capital-rich states may mechanize more
than labor-rich states \citep{gartzke2001democracy}. Lastly,
ideology \citep{van1984cult} and culture \citep[e.g.][]{pollack2004arabs} are
hard-to-measure but perhaps influential. Consider Ireland: between the 1920s and
1940s, it expended a large portion of its defense budget on building
a small number of tanks. They were wholly insufficient to repel a British 
invasion (their ostensible purpose), and
Ireland would have been better-served to pursue a low-mechanization guerrilla
defense strategy, but tanks were seen as a prestigious signifier of a ``real''
professional military, an important factor for the young Irish state
\citep{farrell1998professionalization,farrell2001transnational}.

\citet{sechser2010army} is the landmark study on military mechanization. They
assemble a longitudinal dataset at the country-year level of the mechanization
of all states between 1979 and 2001, and conduct regression analysis to identify
the effect of covariates on mechanization. They conclude that ``choices about
mechanization are strongly associated with a state's security environment.''
The more mechanized a state's geographic neighbors and enemies, the higher the state's 
own mechanization; also, states
learn from defeat in insurgency by subsequently decreasing their mechanization.

@@

The findings of \citet{sechser2010army} are ripe for agentization. The
conclusions drawn implicitly claim
that states are autonomous agents; that they have
cognition regarding how to perceive the outside world
and how to react to it; that spatiality and
neighborhood matter; that inter-agent relationships matter; and that
agents learn from experience. Agent-based modeling is uniquely well-suited to
represent such phenomena \citep{gilbert2005simulation,miller2009complex}.

>>>>

	Military mechanization is the degree to which a state's army is comprised of
	ground combat vehicles as opposed to infantry. Mechanization is both a 
	theoretically important topic in international relations scholarship and a practical 
	concern for policymakers. \citet{sechser2010army} influentially argue that
	inter-state dynamics (e.g., conflict) better explain mechanization choices than do
	intra-state factors (e.g., culture). This paper demonstrates that
	the multivariate statistical approach employed in \citet{sechser2010army} is
	inadequate and that agent-based modeling is needed to capture
	international security competition as a complex system. Candidate
	agent-based models are presented.

@@

Things they don't capture: selection into war and deterrence, gdp / capita to
political structure, institutional culture and choice, network effects and
emulation

@@

Multicollinearity - everything correlated with each other.

@@

This paper presents an agent-based model of military mechanization. It
reviews the academic literature on military mechanization. It discusses why
orthodox modeling strategies fail to capture international security competition
as a complex system and why agent-based modeling is needed. In support of these
claims, it ``agentizes'' the leading model of military mechanization and shows that the
results of doing cast doubt on it. Finally, it presents a
novel agent-based model that better represents military mechanization.

In summary, the orthodox view of mechanization is that it is a
reflexive process: states choose their mechanization posture largely based on
that neighbors, enemies, and recent military experiences (e.g. defeat in
counter-insurgency). This paper argues
that this view---based on traditional longitudinal data analysis---is ripe for
agentization. It implicitly argues for autonomous agents (states) that have
cognition regarding mechanization that takes into account other agents (e.g.
enemies), spatiality (e.g. neighbors), and learning from experience. This paper
builds such a model and validates it against the empirical record, using the
National Mechanization Index of \citet{sechser2010army}. It finds that the 
orthodox view of mechanization does not stand up to agentization. A positive feedback loop 
causes states to
mechanize much faster and to a greater degree than occurs in the real-world.
This implies the existence of some countervailing factor (possibly a negative
feedback loop) that mitigates the extent of mechanization. Candidates for what
this could be mostly center on ``internal'' factors---e.g., culture, economy,
and so on---as opposed to the ``external'', security environment-centric
orthodox view.

\subsection{Background}

States differ in the composition of their ground forces.\footnote{States may have multiple organizations that
conduct ground combat operations: e.g., the United States operates both an Army
and a Marine Corps as independent services. Henceforth, this paper uses the
terms ``army'' and ``ground forces'' interchangeably.} One notable axis of comparison 
is \textit{mechanization}: the degree to which an army consists of infantry 
vs. combat vehicles. Treating mechanization as a continuous variable, 
one extreme is an army consisting exclusively of dismounted, small 
arms-wielding infantry units. The other extreme is an army consisting 
solely of tanks, armored personnel carriers, and the like.
No modern state fully embodies either extreme, but the 1970s-era Vietcong are
a good example of a low-mechanization army and the 1980s-era Israeli army of
high mechanization.

Mechanization is important for multiple academic literatures. A
fundamental debate in international relations (IR) is whether state
behavior is driven by structural insecurity and 
competition \citep{waltz2001man, mearsheimer2001tragedy}; or, whether other
factors such as domestic political structure \citep{russett1994grasping} and
socially constructed beliefs \citep{wendt1992anarchy} can point states to more peaceful directions. 
The extent to which patterns of mechanization do or do not correlate with
\textit{realpolitik} preparation for war informs this debate.
The security studies community is similarly split on the question of military
change. Some view militaries (particularly large ones) as change-averse, 
inertia-driven bureaucracies that can only be bestirred by major events (e.g.,
war or political reform)
\citep{posen1986sources,zegart2000flawed,locher2004victory,halperin2007bureaucratic}. 
Others contend that militaries can and do radically self-transform in both composition and doctrine
\citep{murray1998military,adamsky2010culture}. The patterns of
mechanization---stable? punctuated by sudden change? constantly evolving?---thus
have strong implications for this debate.

Mechanization choices also have more visceral impact.
\citet{lyall2009rage} find that highly mechanized militaries are less effective
at counterinsurgency, while \citet{biddle2004military} argues that
mechanization-enabled mobility aids conventional warfighting. It thus
pays to have an army that is designed to win the kinds of conflicts that a state is 
likely to face. Incorrect forecasting and choosing the wrong mechanization
posture costs blood and treasure. The experience of the U.S. military in the early
2000s, when it shifted under duress from a highly mechanized Cold War-legacy force to a less
mechanized counter-insurgency one, provides a vivid example. A quote from then-Secretary of Defense 
Donald Rumsfeld emphasizes the point:
``you go to war with the army you have, not the army you might wish to have'' \citep{schmitt_2004}.

There are many candidates for the causes of mechanization. Inter-state
competition and war is one. States may choose choose mechanization policies 
that they believe will allow them to prevail against potential adversaries, or to send a
signal of deterrence, or to learn from the perceived mistakes of the last war
\citep{mearsheimer1983conventional,huth1988extended,murray2011military}. 
Economic endowments may play a role: e.g., capital-rich states may mechanize more
than labor-rich states \citep{gartzke2001democracy}. Domestic political
institutions may matter: e.g., autocracies may be more willing than democracies to adopt
low-mechanization ``human wave'' tactics
\citep{quinlivan1999coup,reiter2002democracies,brooks2008shaping,talmadge2015dictator}. 
Ideology \citep{van1984cult} and national culture \citep[e.g.][]{pollack2004arabs} are
hard-to-measure but perhaps influential. Lastly,
organizational culture can play a role \citep{builder1989masks}. Militaries may pursue a
certain mechanization posture because it aligns with their institutional
objectives, which may or may not be coincident with rational strategy. 
Consider Ireland: between the 1920s and 1940s, it seriously considered expending a 
large portion of its defense budget on building a relatively small number of tanks. 
They would have been wholly insufficient to repel a British invasion (their ostensible purpose), 
and Ireland would have been better-served to pursue a low-mechanization guerrilla defense strategy, 
but tanks were seen as a prestigious signifier of a ``real'' professional
military and hence desirable
\citep{farrell1998professionalization,farrell2001transnational}.

\citet{sechser2010army} is the landmark study on military mechanization. They
assemble a longitudinal dataset at the country-year level of the mechanization
of all states between 1979 and 2001. They conduct regression analysis to identify
the effect of covariates on mechanization. They conclude that out of the
aforementioned potential causes---security environment, economy, domestic
politics, and ideology / culture---that the first is most important: ``choices about
mechanization are strongly associated with a state's security environment.''

@@

There is reason to believe that the orthodox multivariate statistical modeling 
approach employed in \citet{sechser2010army} is insufficient. Several factors point towards 
agent-based modeling being a more appropriate choice.

One, from a stat

are ripe for agentization. The
conclusions drawn implicitly claim
that states are autonomous agents; that they have
cognition regarding how to perceive the outside world
and how to react to it; that spatiality and
neighborhood matter; that inter-agent relationships matter; and that
agents learn from experience. Agent-based modeling is uniquely well-suited to
represent such phenomena \citep{gilbert2005simulation,miller2009complex}.
